---
title: "SDM_using_dismo_raster: Prediction"
author: "Hoang Le Xuan"
date: "2024-03-08"
output: html_document
---

Reference:https://rspatial.org/

```{r}
#install.packages(c('raster', 'rgdal', 'dismo', 'rJava'))
library(dismo)
library(maptools)
data(wrld_simpl)

library(dismo)
sdmdata <- readRDS("sdm.Rds")
presvals <- readRDS("pvals.Rds")

sdmdata
```

```{r}
m1 <- glm(pb ~ ., data=sdmdata)
class(m1)
## [1] "glm" "lm"
summary(m1)

#Call:
#glm(formula = pb ~ ., data = sdmdata)
#
#Coefficients: (1 not defined because of singularities)
#              Estimate Std. Error t value Pr(>|t|)    
#(Intercept)  3.931e-01  1.257e-01   3.127 0.001807 ** 
#bclim1      -6.570e-03  3.730e-03  -1.761 0.078457 .  
#bclim10      2.371e-03  6.628e-03   0.358 0.720581    
#bclim11      9.426e-03  5.395e-03   1.747 0.080834 .  
#bclim12     -1.958e-04  7.753e-05  -2.526 0.011663 *  
#bclim13     -9.956e-04  5.347e-04  -1.862 0.062808 .  
#bclim14     -3.258e-03  1.807e-03  -1.802 0.071712 .  
#bclim15     -2.460e-03  5.394e-04  -4.561 5.57e-06 ***
#bclim16      9.743e-04  2.514e-04   3.876 0.000111 ***
#bclim17      1.082e-03  6.342e-04   1.706 0.088219 .  
#bclim18      3.719e-05  9.063e-05   0.410 0.681591    
#bclim19     -7.470e-05  6.750e-05  -1.107 0.268624    
#bclim2       7.806e-03  1.597e-03   4.889 1.14e-06 ***
#bclim3       3.407e-03  1.553e-03   2.194 0.028393 *  
#bclim4       2.622e-04  1.364e-04   1.922 0.054844 .  
#bclim5      -1.170e-02  1.786e-03  -6.554 8.02e-11 ***
#bclim6       4.035e-03  1.777e-03   2.271 0.023327 *  
#bclim7              NA         NA      NA       NA    
#bclim8       3.100e-03  4.728e-04   6.556 7.88e-11 ***
#bclim9       6.681e-04  3.862e-04   1.730 0.083855 .  
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
#
#(Dispersion parameter for gaussian family taken to be 0.1259178)

#    Null deviance: 314.54  on 1347  degrees of freedom
#Residual deviance: 167.34  on 1329  degrees of freedom
#  (1 observation deleted due to missingness)
#AIC: 1053.1

#Number of Fisher Scoring iterations: 2

```

Bioclim uses presence data, so we use ‘presvals’ instead of ‘sdmdata’.
```{r}
bc <- bioclim(presvals[,colnames(presvals)])
bc <- bioclim(presvals[,c("bclim12", "bclim15",  "bclim16",  "bclim2",  "bclim3", "bclim5", "bclim8", "bclim6" )])

class(bc)
#[1] "Bioclim"
#attr(,"package")
#[1] "dismo"

bc

pairs(bc)
```

## Model prediction

response plots for each variable, with the other variables at their median value.
```{r}
response(bc)

```

The purpose is SDM is to create a map of suitability scores. We create by providing the predict function with a Raster* object and a model object.

```{r}
predictors <- stack(list.files(path="data/bioclim",  pattern='asc',  full.names=TRUE ))
names(predictors)
## [ [1] "bclim1"  "bclim10" "bclim11" "bclim12" "bclim13" "bclim14" "bclim15" "bclim16" "bclim17" "bclim18"
## [11] "bclim19" "bclim2"  "bclim3"  "bclim4"  "bclim5"  "bclim6"  "bclim7"  "bclim8"  "bclim9"
p <- predict(predictors, m1)
```

```{r}
plot(p)

```

## Model evaluation:

illustrate the computation of the correlation coefficient and AUC with two random variables. p (presence) has higher values, and represents the predicted value for 50 known cases (locations) where the species is present, and a (absence) has lower values, and represents the predicted value for 50 known cases (locations) where the species is absent.

```{r}
p <- rnorm(50, mean=0.7, sd=0.3)
a <- rnorm(50, mean=0.4, sd=0.4)
par(mfrow=c(1, 2))
plot(sort(p), col='red', pch=21)
points(sort(a), col='blue', pch=24)
legend(1, 0.95 * max(a,p), c('presence', 'absence'),
          pch=c(21,24), col=c('red', 'blue'))
comb <- c(p,a)
group <- c(rep('presence', length(p)), rep('absence', length(a)))
boxplot(comb~group, col=c('blue', 'red'))
```
Compute the correlation coefficient and the AUC

```{r}
group = c(rep(1, length(p)), rep(0, length(a)))
cor.test(comb, group)$estimate
##       cor
## 0.4151511
mv <- wilcox.test(p,a)
auc <- as.numeric(mv$statistic) / (length(p) * length(a))
auc
```
```{r}
e <- evaluate(p=p, a=a)
class(e)
## [1] "ModelEvaluation"
## attr(,"package")
## [1] "dismo"
e

## class          : ModelEvaluation 
## n presences    : 50 
## n absences     : 50 
## AUC            : 0.6592 
## cor            : 0.2594386 
## max TPR+TNR at : 0.4662509 
par(mfrow=c(1, 2))
density(e)
boxplot(e, col=c('blue', 'red'))
```

Back to some real data, presence-only in this case. We’ll divide the data in two random sets, one for training a Bioclim model, and one for evaluating the model.

```{r}
samp <- sample(nrow(sdmdata), round(0.75 * nrow(sdmdata)))
traindata <- sdmdata[samp,]
traindata <- traindata[traindata[,1] == 1, 2:9]
testdata <- sdmdata[-samp,]
bc <- bioclim(traindata)
e <- evaluate(testdata[testdata==1,], testdata[testdata==0,], bc)
e
##class          : ModelEvaluation 
##n presences    : 220 
##n absences     : 117 
##AUC            : 0.8366939 
##cor            : 0.4202167 
##max TPR+TNR at : 0.05244777 
plot(e, 'ROC')
```
## k-fold data partitioning
first create presence and background data
```{r}
pres <- sdmdata[sdmdata[,1] == 1, 2:9]
back <- sdmdata[sdmdata[,1] == 0, 2:9]

#k <- 5
group <- kfold(pres, k)
group[1:10]
##  [1] 5 2 2 5 4 3 1 3 5 4
unique(group)
## [1]  5 2 4 3 1

##fit and test our model five times. In each run, the records corresponding to one of the five groups is only used to evaluate the model, while the other four groups are only used to fit the model. The results are stored in a list called ‘e’.
e <- list()
for (i in 1:k) {
    train <- pres[group != i,]
    test <- pres[group == i,]
    bc <- bioclim(train)
    e[[i]] <- evaluate(p=test, a=back, bc)
}
```


```{r}
auc <- sapply(e, function(x){x@auc})
auc
## [1] 0.7739565 0.8291739 0.7931250 0.7588696 0.7660435
mean(auc)
## [1] 0.7842337
sapply( e, function(x){ threshold(x)['spec_sens'] } )
```

```{r}
e <- list()
for (i in 1:k) {
    train <- pres[group != i,]
    test <- pres[group == i,]
    bc <- bioclim(train)
    e[[i]] <- evaluate(p=test, a=back, bc)
}
```

```{r}
auc <- sapply(e, function(x){x@auc})
auc
## [1] 0.7739565 0.8291739 0.7931250 0.7588696 0.7660435
mean(auc)
## [1] 0.7842337
sapply( e, function(x){ threshold(x)['spec_sens'] } )
```



```{r}
canetoad <- read.table("data/ctsel.csv", header = T, sep=',')
canetoad <- canetoad[,-1]

#omit this
#canetoad <- canetoad[,-1]
presvals <- extract(predictors, canetoad)
set.seed(0)
backgr <- randomPoints(predictors, 500)
nr <- nrow(canetoad)
s <- sample(nr, 0.25 * nr)
pres_train <- canetoad[-s, ]
pres_test <- canetoad[s, ]
nr <- nrow(backgr)
set.seed(9)
s <- sample(nr, 0.25 * nr)
back_train <- backgr[-s, ]
back_test <- backgr[s, ]
```

```{r}
sb <- ssb(pres_test, back_test, pres_train)
sb[,1] / sb[,2]
```

```{r}
i <- pwdSample(pres_test, back_test, pres_train, n=1, tr=0.1)
pres_test_pwd <- pres_test[!is.na(i[,1]), ]
back_test_pwd <- back_test[na.omit(as.vector(i)), ]
sb2 <- ssb(pres_test_pwd, back_test_pwd, pres_train)
sb2[1]/ sb2[2]
## [1] 1.004106
```

```{r}
bc <- bioclim(predictors, pres_train)
evaluate(bc, p=pres_test, a=back_test, x=predictors)
## class          : ModelEvaluation
## n presences    : 29
## n absences     : 125
## AUC            : 0.757931
## cor            : 0.2777298
## max TPR+TNR at : 0.03438276
evaluate(bc, p=pres_test_pwd, a=back_test_pwd, x=predictors)
```



